---
title: "ESM 244 Lab 1"
author: "Rachel Behm"
date: "January 10, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Objectives:
-basic data wrangling review
-intro to spatial data in r with sf
-intro to shiny apps


**ctr+alt+i = shortcut to insert r script**

1. Packages
```{r, include = FALSE}
library(tidyverse)
library(sf)
library(leaflet)
library(tmap)
```

2. Review of reading, exploring, wrangling, and visualizing data
a. read in data
```{r, include= FALSE}
wine<-read_csv("wine_data.csv")

#if want to look at do View(wine) in the console
```

b.wrangling
want to find wine that is most bang for your buck

**r+shift+m= shortcut for pipe operator**
```{r}
wine_new<- wine %>% 
  select(country, province, winery, region_1, points,price) %>% 
  rename(state = province) %>% 
  filter(state =="California" | state == "Oregon" |state == "Washington") %>% 
  mutate(ppd = points/price) %>% 
  arrange(-ppd)

#select chooses desired columns and/or reorders columns
# one equal =ets a value equal to something == looks for a match
#ppd= points per dollar. want higher ppd
#arrange- sort by decreasing ppd (1st is the highest)
```


c. use group_by and summarize to find summary statistics by group
want to look at state comparison easily
```{r}
wine_summary <- wine_new %>% 
  group_by(state) %>%
  summarize(
    mean_ppd = mean(ppd, na.rm = TRUE),
    max_ppd = max(ppd, na.rm=TRUE)
  )

#washington has higher avg ppd than other two
#max is actually in CA
#if use another var in group_by will seperate out those characters seperately
```

d.gg graphics
```{r}
wine_hists <- ggplot(wine_new, aes(x= ppd)) +
  geom_histogram(aes(fill=state)) +
  facet_wrap(~state, scale = "free")

wine_hists
#facet_wrap = facet_grid
#having the wrap default set to not free is good. ppl could be misled make sure to be clear if change
```
**POINTS= COLOR. BAR= FILL** 

scatterplot of price vs points with color by state and linear model trendline on top
```{r}
wine_points <- ggplot(wine_new, aes(x= price, y = points))+
  geom_point(aes(color=state), alpha =0.5) +
  geom_smooth(color="black", method = "lm", se = FALSE) +
  scale_color_manual(values = c("royalblue","mediumorchid1", "orange")) +
  facet_wrap(~state)+
  theme_light()

wine_points
#scale_color_ can manually change color how u want
# if data is hard to see, separate the views
#adding trendline is literally so easy via geom_smooth woot woot
#geom_smooth() = make a line that is
#color= "black" = the color black
#,method = "lm" = and is of the linear model trendlilne
#,se = FALSE = but dont show the standard error
 
```
 
 2. california county spatial data via sf
```{r}
ca_counties <- st_read(dsn = ".", layer = "CA_Counties_TIGER2016" )

#since not csv file have to use st_read
#dsn = destination working directory = "." we already there lol

#plot(ca_counties)


#shows everything but we only care about income
#sf uses sticky geometry so you dont have to worry about dealing with that thank you based god

ca_land <- ca_counties %>%
  select(NAME, ALAND)

ca_pop_inc <- read_csv("ca_pop_inc.csv") %>%
  rename(NAME = COUNTY)
#now we want to combine with the spatial data- what name do the two have in common? they have names but the title doesnt match, we have to fix that. everything has to match even the values

ca_df <- full_join(ca_land, ca_pop_inc) %>% 
  select(NAME, MedFamilyIncome)

#full_join is the safest way because doesnt get rid of anything
#R will match the column with the same name and add all of the new columns to make new df
#now we wanna make a map of the data
#spatial info is still there even after select woooo

```
 
d. use geom_sf to make maps in ggplot
```{r}
ca_income <- ggplot(ca_df) +
  geom_sf(aes(fill = MedFamilyIncome))

ca_income
#dont need x and y axis since using spatial data

#leaflet(ca_df) %>% 
  #addPolygons()
#doesnt work cause datum doesnt match



ca_df_transform <- st_transform(ca_df, crs = 4326)
leaflet(ca_df_transform) %>% 
  addTiles() %>%
  addPolygons()

#makes interactive map of polygons
```

